Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-07-24 manuscript.tex
refer to Fig 5 and add high freq Va eqn

There are several points worth noting about this figure. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Ref. \cite{Dendy1994} (also
Fig. 9 Ref. \cite{Dendy1995}) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign, which pushes the real
frequency from one side of the harmonic to the other. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.

>>>>

There are several points worth noting about Fig. \ref{fig:2D_F12_physicalpitch}.
First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Ref. \cite{Dendy1994} (also
Fig. 9 Ref. \cite{Dendy1995}) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign, which pushes the real
frequency from one side of the harmonic to the other. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation.

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and
propagation angles that are close to perpendicular.

Figure \ref{fig:2D_real_zeropitch}
shows the smoothed real part of the solution for the case of zero pitch angle
cosine and is some normalised by
the high frequency expression for the fast Alfv{\'e}n wave set out in Eq.
25 of Ref. \cite{Dendy1994} (also Eq. 11 of Ref. \cite{McClements1996}),

\begin{equation}
  \omega^2 = \frac{V_A^2}{2}\left[
    k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}
  + \sqrt{
    \left(k^2 + k_\parallel^2 + k^2k_\parallel^2 \frac{V_A^2}{\Omega_i^2}\right)^2
    - 4k^2 k_\parallel^2
}
  \right],
\end{equation}

\noindent where $k^2 = k_\parallel^2 + k_\perp^2$. The real part of the
solutions for the case of non-zero pitch angle cosine is not discernably
different.
